---
title: "Predicting Weight-Lifting Quality"
author: "Aaron Ferrucci"
date: "October 17, 2015"
output: 
  html_document:
    fig_height: 4
    fig_width: 6
references:
- id: Ugulino2012
  title: "Wearable Computing: Accelerometers' Data Classification of Body Postures and Movements"
  author:
  - family: Ugulino, W.; Cardador, D.; Vega, K.; Velloso, E.; Milidiu, R.; Fuks, H.
  container-title: Proceedings of 21st Brazilian Symposium on Artificial Intelligence
  URL: 'http://groupware.les.inf.puc-rio.br/har'
---


``` {r echo=FALSE, include=F, messages=F, warnings=F, results="hide"}
# Load libraries, data
set.seed(12345)
library(caret)
library(randomForest)
```

## Overview
Accelometer devices make it possible to collect data from subjects
performing weight-lifting exercises. It's possible that user technique
during such exercises may be significant - indicating risk of injury, or
low rate of improvement. Quantifying the _quality_ of exercise; whether
in real time, or in post-processing could be a valuable training aid.

This project takes as input a set of data gathered during weight-lifting
exercises, and attempts to discern, via machine learning classification
modeling, whether the exercises were done correctly or incorrectly.

If a reasonable predictive model can be built, it could be implemented
as a fitness product on a mobile device. Because mobile devices are typically
underpowered in terms of compute performances, finding a model with lower
compute requirements is desirable.

## Preliminaries
The test and training data sets are provided as files in .csv format. First,
I'll load in the training data set and do some basic exploratory analysis.

``` {r echo=T}
train_all <- read.csv("pml-training.csv")
```

The data set consists of `r nrow(train_all)` observations of `r ncol(train_all)`
variables.  As explained in [@Ugulino2012], the observations are of 6
individuals performing a weight-lifting exercise correctly, and then
incorrectly in a number of ways. Correctness or not is encoded in the
`classe` variable of the dataset, which is a factor variable of 5 levels, with
the following interpretation:

* A: exactly according to the specification
* B: throwing the elbows to the front
* C: lifting the dumbbell only halfway
* D: lowering the dumbbell only halfway
* E: throwing the hips to the front 

A contingency table by subject and class shows how many of each type of measurement were made, for each individual:
``` {r}
table(train_all$classe, train_all$user_name)
```

## Reducing Covariates
To improve scalability of any classification algorithm, it's worth trying to
reduce the data size to only those measurements which matter. 

The dataset contains a large number measurements which are NA or empty string
(""):

``` {r}
nas <- apply(train_all, 2, function(x) sum(ifelse(is.na(x), 1, 1 * (x == ""))))
t <- table(nas)
t
```

When a measurement contains any missing values, it contains a 
large number of them (`r rownames(t)[2]` out of `r nrow(train_all)` 
total measurements).  Looking closer, I
see that variables which have mostly missing values have actual values when 
the variable `r "new_window"` is "yes". The names of those variables 
indicate that they are summary statistics (min, max, avg, etc.) - in other 
words, they are
transformed tidy or Level 2 covariates. In the interest of letting
the machine learning algorithm discover the necessary variables, I will 
discard all these transformed covariates.

``` {r}
train_all <- train_all[, nas == 0]
```

Next, I'll find and eliminate any covariates which have little impact on
the outcome (via `nearZeroVar`):

``` {r}
nzv <- nearZeroVar(train_all, saveMetrics=T)
train_all <- train_all[,!nzv$nzv]
```

Finally, I'll manually discard a few covariates based on "common sense". 
For example,
the `X` value is a simple index over all of the data; there are also a 
few timestamp values that are unlikely to be predictors for correct vs. 
incorrect weightlifting technique.

``` {r}
train_all <- train_all[, -grep("timestamp|X|num_window", names(train_all))]
```

I've reduced the number of covariates to `r ncol(train_all)`, which is a big
improvement over the original data size.

## Preparing for Cross-Validation
I will separate the training set into train and validation subsets, and use 
the validation subset error as my out-of-sample error estimate. A good rule
of thumb is to separate a data set into 60% training, 20% validation and 20%
test. In this case, the test data has already been separated out. By the rule 
of thumb, the training and validation data are in a 3:1 ratio, so I'll
maintain that for my training and valaidation data with `p=0.75`.
``` {r}
dp <- createDataPartition(y=train_all$classe, p=0.75, list=F) 
train <- train_all[dp,]
validate <- train_all[-dp,]
```

## Random Forest - Initial Model
Using the separated training data, I'll train a random forest, using all the
default options. Despite the rumors, creation of the model takes only 
a few minutes, even on my ancient laptop. However, an attempt to build the
same model with caret's `train` function took much longer - so I avoided 
that.
``` {r cache=TRUE}
rf <- randomForest(classe ~ ., data=train)
rf
```

### Examining the Initial Model: Estimate Out-of-Sample Error
### OOB error
One of the figures of merit reported for the random forest model is the
OOB estimate. In theory, this estimate is an accurate estimate of the
out-of-sample error.  For this model, the OOB estimate was very low, 
indicating a good fit without danger of overfitting.

### Out-of-Sample error: Home brew
As a cross-check, I'll implement my own out-of-sample estimate, using 
the reserved validation data. The concept here is that since I don't have 
access to the test data, I'll estimate the error on the test set using 
training data which was separated out before building the model.

How does the model perform on the validation data?
``` {r}
modelError <- function(fit, data) {
  missClass <- sum(predict(fit, data) == data$classe)
  err <- missClass / nrow(data)
  return(100 - err * 100)
}

me <- modelError(rf, validate)
me
```

The out-of-sample error estimate is not dissimilar to the random forest's
OOB estimate.

### Examining the Initial Model: Error vs. Number of Trees
``` {r}
plot(rf)
legend("topright", legend=colnames(rf$err.rate), lty=seq(ncol(rf$err.rate)), col=seq(ncol(rf$err.rate)))
```

After only 50 trees or so, the error rate has settled nicely.
`randomForest()`'s default number of trees (500) is probably overkill.

### Examining the Initial Model: Variable Importance
Now that the model is trained, it's interesting to have a look at variable
importance. Since the model appears to be very accurate, could we trim this 
model, for better scalability, without sacrificing too much accuracy?
``` {r}
order <- order(-rf$importance)
plot(
  rf$importance[order],
  main="Variable Importance",
  ylab = "Mean Gini Decrease",
  xlab=""
)
```

The out-of-sample error estimate is low; the plot of 
variable importance shows that a few variables explain most of the 
prediction. Taken together, these facts make it seem likely that an 
accurate model can be built using much fewer of the variables.

## Random Forest - Minimized Model
I'll try training a new "minimized" random forest, using only the top 15 
variables, and using only 100 trees.
``` {r cache=T}
impNames <- rownames(rf$importance)[order]
minimal <- c(impNames[1:15], "classe")
dp_minimal <- createDataPartition(y=train_all$classe, p=0.75, list=F) 
train_minimal <- train_all[dp_minimal,minimal]
validate_minimal <- train_all[-dp_minimal,minimal]
rf_minimal <- randomForest(classe ~ ., data=train_minimal, ntree=100)
rf_minimal
```

### Minimized Model: Out-of-Sample Error Estimate
The minimized model has a slightly reduced OOB estimate, but it's still quite
good. Again using the validation data set, I  estimate the
out-of-sample error:

``` {r}
me_minimal <- modelError(rf_minimal, validate_minimal)
me_minimal
```

Once again, a similar figure is reported for the out-of-sample error estimate.

### Minimized Model: Error vs. Number of Trees
``` {r}
plot(rf_minimal)
legend("topright", legend=colnames(rf_minimal$err.rate), lty=seq(ncol(rf_minimal$err.rate)),
col=seq(ncol(rf_minimal$err.rate)))
```

Again, after 50 trees or so, the error rate has settled nicely. So this
"minimal" model should predict with reasonable accuracy.

## Model Final Test
As a final test, I'll compare results between the initial and the minimal
models, on the test data. This will be a relative test, comparing the results
of the two models. I won't feed back any info garnered here
back into the models; therefore I think this is an allowed use of test data.
``` {r}
test <- read.csv("pml-testing.csv")
answers <- predict(rf, test)
answers_minimal <- predict(rf_minimal, test)
all(answers == answers_minimal)
```

The initial and the minimized model predict the same results on the test
data set, so at the least, it appears that the minimal model could provide
reasonable results with better scalability.

## Conclusions
* A complex and a simple random forest model were built; both achieved 
 reasonable OOB estimates
* Out-of-sample estimates were calculated using cross-validation; the results
 were similar to the OOB estimates.
* The models predicted identical results on the reserved test data set.
* The minimal model may be suited to implementation on a low-powered mobile
 device - further profiling and optimization are required.

``` {r eval=FALSE, echo=FALSE}
test <- read.csv("pml-testing.csv")
answers <- predict(rf, test)
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n) {
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}
answers
pml_write_files(answers)
```
## Appendix - Reproducibility
For reproducibility, here's the output of sessionInfo():
``` {r}
sessionInfo()
```

## References

